---
title: "Solutions 7"
callout-icon: false
code-fold: show
---

```{r}
#| label: "Set up"
#| message: false

library(tidyverse)
library(infer)
library(gssr)

theme_set(theme_light(base_family = "Optima"))
```

## Contingency Tables

```{r}
#| message: false
gss18 <- gss_get_yr(2018) 

d <- gss18 |> 
  select(attend, polviews) |> 
  haven::zap_missing() |> 
  haven::zap_labels() |> 
  mutate(weekly = if_else(attend >= 7, 1L, 0L),
         conservative = if_else(polviews >= 5, 1L, 0L)) |> 
  drop_na() 

crosstab1 <- table(conservative = d$conservative, weekly = d$weekly)
crosstab1
```

### Exercise

It might be easier to modify the table so that we look at the margins:

```{r}
addmargins(crosstab1)
```

I will provide answers using (1) the contingency table, (2) base R, and (3) tidyverse

::: callout-note
What is the probability that a person attends religious services weekly?

I'm asking for $\Pr(W)$
:::

```{r}
## contingency table
497 / 2235
## base R
mean(d$weekly)
## tidyverse
d |> 
  summarize(p = mean(weekly))
```

::: callout-note
What is the probability that a person *does not* attend religious services weekly?

I'm asking for $\Pr(W^\complement)$
:::

```{r}
## contingency table
1738 / 2235
## base R
mean(d$weekly == 0)
## tidyverse
d |> 
  summarize(p = mean(weekly == 0))
```

::: callout-note
What is the probability that a person is conservative, given that they attend religious services weekly?

I'm asking for $\Pr(C \mid W)$
:::

```{r}
## contingency table
240 / 497
## base R
mean(d$conservative[d$weekly == 1])
## tidyverse
d |> 
  filter(weekly == 1) |> 
  summarize(p = mean(conservative))
```

::: callout-note
$\Pr(W \mid C)$
:::

```{r}
## contingency table
240 / 735
## base R
mean(d$weekly[d$conservative == 1])
## tidyverse
d |> 
  filter(conservative == 1) |> 
  summarize(p = mean(weekly))
```

::: callout-note
$\Pr(C \mid W^\complement)$
:::

```{r}
## contingency table
495 / 1738
## base R
mean(d$conservative[d$weekly == 0])
## tidyverse
d |> 
  filter(weekly == 0) |> 
  summarize(p = mean(conservative))
```

::: callout-note
$\Pr(W \text{ and } C)$
:::

```{r}
## contingency table
240 / 2235
## base R
mean(d$weekly & d$conservative) ## this does implicit coercion, see Week 1!
## tidyverse
d |> 
  count(weekly, conservative) |> 
  mutate(p = n / sum(n))
```

::: callout-note
$\Pr(W^\complement \text{ and } C^\complement)$
:::

```{r}
## contingency table
1243 / 2235
## base R
mean(d$weekly == 0 & d$conservative == 0)
## tidyverse
d |> 
  count(weekly, conservative) |> 
  mutate(p = n / sum(n)) |> 
  filter(weekly == 0, conservative == 0)
```

::: callout-note
Finally, if we *assume* that $W$ and $C$ are independent:

-   What *would* be the joint probability of $W$ and $C$ ?
:::

```{r}
(735 / 2235) * (497 / 2235)
```

::: callout-tip
Hint: These are some useful functions for working with contingency tables.

```{r}
addmargins(crosstab1)
rowSums(crosstab1)
colSums(crosstab1)
prop.table(crosstab1)
prop.table(crosstab1, margin = 1) ## row percentages
prop.table(crosstab1, margin = 2) ## col percentages
```
:::

### Exercise

Here I chose the `cappun` variable, which asks *Do you favor or oppose the death penalty for persons convicted of murder?*

```{r}
d <- gss18 |> 
  select(attend, polviews, cappun) |> 
  haven::zap_missing() |> 
  haven::zap_labels() |> 
  mutate(
    weekly = if_else(attend >= 7, 1L, 0L),
    conservative = if_else(polviews >= 5, 1L, 0L),
    cappun = if_else(cappun == 1, 1L, 0L)
    ) |> 
  drop_na() 

crosstab2 <- table(conservative = d$conservative, capital_punishment = d$cappun)
addmargins(crosstab2)
```

Marginal probabilities:

```{r}
table(capital_punishment = d$cappun) |> 
  prop.table()
```

-   The probability that a person favors capital punishment is 0.63

-   The probability that a person does *not* favor capital punishment is 0.37

Now look at this:

```{r}
prop.table(crosstab2, margin = 1)
```

|                                                 |                                                                       |
|-------------------------------------------------|-----------------------------------------------------------------------|
| $\Pr(CP \mid C) = 0.75$                         | The probability that a conservative person favors capital punishment. |
| $\Pr(CP^\complement \mid C) = 0.25$             | The probability that a conservative person opposes capital punishment |
| $\Pr(CP \mid C^\complement) = 0.57$             | The probability that a non-conservative favors capital punishment.    |
| $\Pr(CP^\complement \mid C^\complement) = 0.42$ | The probability that a non-conservative opposes capital punishment.   |

Now look at this:

```{r}
prop.table(crosstab2, margin = 2)
```

|                                                 |                                                                          |
|-------------------------------------------------|--------------------------------------------------------------------------|
| $\Pr(C \mid CP) = 0.4$                          | The probability that a capital punishment supporter is conservative.     |
| $\Pr(C \mid CP^\complement) = 0.22$             | The probability that a capital punishment opposer is conservative.       |
| $\Pr(C^\complement \mid CP) = 0.6$              | The probability that a capital punishment supporter is non-conservative. |
| $\Pr(C^\complement \mid CP^\complement) = 0.77$ | The probability that a capital punishment opposer is non-conservative.   |

Now look at this:

```{r}
prop.table(crosstab2)
```

|                                                         |                                                                                      |
|---------------------------------------------------------|--------------------------------------------------------------------------------------|
| $\Pr(C \text{ and } CP) = 0.25$                         | The probability that a person is conservative *and* favors capital punishment.       |
| $\Pr(C \text{ and } CP^\complement) = 0.08$             | The probability that a person is conservative *and* opposes capital punishment.      |
| $\Pr(C^\complement \text{ and } CP) = 0.38$             | The probability that a person is non-conservative *and* favors capital punishment.   |
| $\Pr(C^\complement \text{ and } CP^\complement) = 0.28$ | The probability that a person is non-conservative *and* opposes capital punishment.  |

Assuming independence:

```{r}
conservative <- table(conservative = d$conservative)
conservative <- conservative / sum(conservative)
cp <- table(capital_punishment = d$cappun)
cp <- cp / sum(cp)
outer(conservative, cp)
```

*Here, each cell is the product of the marginal probabilities. You probably calculated these probabilities differently. That's OK.*

## Four Ways to Make Sense of Dependence

### Exercise

::: callout-note
Take another look at the `weekly` vs `conservative` contingency table.

Focus on these two probabilities:

$$
\begin{align}
\Pr(\text{weekly} \mid \text{conservative}) &&\text{and} && \Pr(\text{weekly} \mid \text{non-conservative})
\end{align}
$$

Compare these probabilities using the following summary statistics:

1.  Difference in Probabilities
2.  Relative Risk Ratio
3.  Odds Ratio
4.  Log Odds Ratio
:::

*Note. Recall that---under independence---these two probabilities would be the same. So there would be (1) zero difference in probabilities; (2) a relative risk ratio of one; (3) an odds ratio of one; and (4) a log odds ratio of zero.*

**Difference in proportions.**

```{r}
prop.table(crosstab1, margin = 1)
0.3265306 - 0.1713333
```

Conservatives are 15 percentage points more likely to attend religious ceremonies weekly.

**Relative Risk Ratio**

```{r}
0.3265306 / 0.1713333
```

Conservatives' probability of attending religious ceremonies weekly is almost 2 times greater.

*Note the asymmetry. Non-conservatives' probability of not attending weekly is 1.2 times greater.*

```{r}
0.8286667 / 0.6734694
```

**Odds ratio**

```{r}
## odds that a conservative attends RC weekly
numerator <- 0.3265306 / (1 - 0.3265306)
numerator
## odds that a non-conservative attends RC weekly
denominator <- 0.1713333 / (1 - 0.1713333)
denominator
## odds ratio
numerator / denominator
```

*Note the symmetry now:*

```{r}
(0.8286667 / (1 - 0.8286667)) / (0.6734694 / (1 - 0.6734694))
```

**Log Odds Ratio**

```{r}
log(numerator / denominator)
log(denominator / numerator)
```

### Exercise

::: callout-note
Last week, in class, we used the `infer` package to draw a "simulation-based bootstrap distribution" for the difference in proportions between $\Pr(W \mid C)$ and $\Pr(W \mid C^\complement)$.

Now we mix things up a little bit.

1.  Draw a sampling distribution for the "relative risk ratio" (Hint: `infer` calls this "ratio of props")
2.  Draw a sampling distribution for the "odds ratio"
:::

**Using `infer`**

```{r}
d <- d |> 
  mutate(
    CP = if_else(cappun == 1, "supports", "opposes"),
    C = if_else(conservative == 1, "conservative", "other"),
    W = if_else(weekly == 1, "weekly", "other")
  )

boot_diff_in_props <- d |> 
  specify(W ~ C, success = "weekly") |> 
  generate(reps = 5e3, type = "bootstrap") |> 
  calculate("diff in props", order = c("conservative", "other"))

boot_diff_in_props |> 
  ggplot(aes(stat)) + 
  geom_histogram(color = "white", boundary = 0.1, binwidth = 0.005) + 
  geom_vline(xintercept = 0, linetype = "dashed") + 
  labs(
    title = "Difference in Proportions", 
    subtitle = "Simulation-Based Bootstrap Distribution"
  )

boot_rrr <- d |> 
  specify(W ~ C, success = "weekly") |> 
  generate(reps = 5e3, type = "bootstrap") |> 
  calculate("ratio of props", order = c("conservative", "other"))

boot_rrr |> 
  ggplot(aes(stat)) + 
  geom_histogram(color = "white", binwidth = 0.05) + 
  geom_vline(xintercept = 1, linetype = "dashed") +
  labs(
    title = "Relative Risk Ratio", 
    subtitle = "Simulation-Based Bootstrap Distribution"
  )

boot_odds_ratio <- d |> 
  specify(W ~ C, success = "weekly") |> 
  generate(reps = 5e3, type = "bootstrap") |> 
  calculate("odds ratio", order = c("conservative", "other"))

boot_odds_ratio |> 
  ggplot(aes(stat)) + 
  geom_histogram(color = "white", binwidth = 0.1) + 
  geom_vline(xintercept = 1, linetype = "dashed") + 
  labs(
    title = "Odds Ratio", 
    subtitle = "Simulation-Based Bootstrap Distribution"
  )
```

*Note. It's easy to draw the sampling distribution for the log odds!*

```{r}
boot_odds_ratio |> 
  mutate(stat = log(stat)) |> 
  ggplot(aes(stat)) + 
  geom_histogram(color = "white", binwidth = 0.05) + 
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    title = "Log Odds Ratio", 
    subtitle = "Simulation-Based Bootstrap Distribution"
  )
```

### Exercise

::: callout-note
Take another look at the `cappun` vs `conservative` contingency table.

Focus on these two probabilities:

$$
\begin{align}
\Pr(\text{supports cappun} \mid \text{conservative}) &&\text{and} && \Pr(\text{supports cappun} \mid \text{non-conservative})
\end{align}
$$

Compare these probabilities using the following summary statistics:

1.  Difference in Probabilities
2.  Relative Risk Ratio
3.  Odds Ratio
4.  Log Odds Ratio
:::

**Difference in Proportions**

```{r}
prop.table(crosstab2, margin = 1)
0.7489240 - 0.5729537
```

Conservatives' support for capital punishment is 17 percentage points higher than non-conservatives'.

**Relative Risk Ratio**

```{r}
0.7489240 / 0.5729537
```

Conservatives' support for capital punishment is 1.3 times greater than non-conservatives'.

**Odds ratio**

```{r}
## odds that a conservative supports CP
numerator <- 0.7489240 / (1 - 0.7489240)
numerator
## odds that a non-conservative supports CP
denominator <- 0.5729537 / (1 - 0.5729537)
denominator
## odds ratio
numerator / denominator
```

The odds that a conservative supports capital punishment is 2 times larger than the odds that non-conservative supports capital punishment.

**Log Odds Ratio**

```{r}
log(numerator / denominator)
```

### Exercise

**Sampling distribution for the relative risk ratio**

```{r}
boot_rrr <- d |> 
  specify(CP ~ C, success = "supports") |> 
  generate(reps = 5e3, type = "bootstrap") |> 
  calculate("ratio of props", order = c("conservative", "other"))

boot_rrr |> 
  infer::visualize() + 
  geom_vline(xintercept = 1, linetype = "dashed")
```

**Sampling distribution for the odds ratio**

```{r}
boot_odds_ratio <- d |> 
  specify(CP ~ C, success = "supports") |> 
  generate(reps = 5e3, type = "bootstrap") |> 
  calculate("odds ratio", order = c("conservative", "other"))

boot_odds_ratio |> 
  infer::visualize() + 
  geom_vline(xintercept = 1, linetype = "dashed")
```

**How to do this *without* using the `infer` package?**

```{r}
out <- replicate(5e3, {
  
  i <- sample(1:nrow(d), replace = TRUE)
  dr <- d[i, ]  ## integer subsetting (see Week 1)
  
  out <- table(
    conservative = dr$conservative, 
    capital_punishment = dr$cappun
  ) |> prop.table(margin = 1)
  
  rrr <- out[2, 2] / out[1, 2]
  odds_ratio <- (out[2, 2] / (1 - out[2, 2])) / (out[1, 2] / (1 - out[1, 2]))
  
  tibble(rrr, odds_ratio)
  
}, simplify = FALSE)

boot <- bind_rows(out) |> 
  rowid_to_column(var = "replicate") |> 
  mutate(log_odds_ratio = log(odds_ratio))

glimpse(boot)

boot |> 
  ggplot(aes(rrr)) +
  geom_histogram(color = "white") +
  geom_vline(xintercept = 1, linetype = "dashed")

boot |> 
  ggplot(aes(odds_ratio)) +
  geom_histogram(color = "white") +
  geom_vline(xintercept = 1, linetype = "dashed")
```
