---
title: "Solutions 5"
callout-icon: false
code-fold: true
---

```{r}
#| label: Set up
#| message: false
library(tidyverse)
theme_set(theme_light(base_family = "Optima"))
```

## Probability

### Exercise

::: callout-note
Suppose that $p = 2/3$.

-   Use `dbinom()` to verify the results in @eq-bernouilli-pmf.

-   Use `pbinom()` to verify the results in @eq-bernouilli-cdf.

-   Use `rbinom()` to generate 100 samples from the Bernouilli distribution with `prob = 2/3`. Plot the results in `ggplot` using `geom_bar`.
:::

`dbinom()`

```{r}
## should be 2/3
dbinom(1, size = 1, prob = 2/3)
## should be 1/3
dbinom(0, size = 1, prob = 2/3) 
## should all be zero
dbinom(c(-2, -1, 2, 3), size = 1, prob = 2/3) 
```

`pbinom()`

```{r}
## should be all zero 
pbinom(c(-3, -2, -1), size = 1, prob = 2/3)
## should be 1/3
pbinom(0, size = 1, prob = 2/3)
## should be 1
pbinom(1, size = 1, prob = 2/3)
## should all be 1
pbinom(c(2, 3, 4, 5), size = 1, prob = 2/3)
```

`rbinom()`

```{r}
x <- rbinom(100, size = 1, prob = 2/3)
glimpse(x)
mean(x)

tibble(x = as_factor(x)) |> 
  ggplot(aes(x)) + 
  geom_bar(width = 1/2)
```

### Exercise

::: callout-note
Suppose we flip a fair coin exactly four times.

Draw the sample space $\Omega$ for this "experiment."
:::

$$
\Omega = \begin{Bmatrix} 
T, T, T, T \\
\hline
H, T, T, T \\ T, H, T, T \\ T, T, H, T \\ T, T, T, H \\
\hline
H, H, T, T \\ H, T, H, T \\ H, T, T, H \\ T, H, H, T \\ T, H, T, H \\ T, T, H, H \\
\hline
T, H, H, H \\ H, T, H, H \\ H, H, T, H \\ H, H, H, T \\
\hline
H, H, H, H
\end{Bmatrix}
$$

<aside>The horizontal lines will prove useful when we turn this sample space into a random variable.</aside>

### Exercise

::: callout-note
Turn the previous sample space into a random variable that counts the number of heads ( $H$ ) in each event.

-   How many possible ways are there to get $X = 0$, $X = 1$, $X = 2$, $X = 3$, and $X = 4$?

-   What is the probability that $X = 2$?

-   Verify this by using the correct R function to calculate the probability that $X = 2$?

-   Use the correct R function to calculate the probability that $X \leq 1$?
:::

$$
\begin{align}
\Omega = \begin{Bmatrix} 
T, T, T, T \\
\hline
H, T, T, T \\ T, H, T, T \\ T, T, H, T \\ T, T, T, H \\
\hline
H, H, T, T \\ H, T, H, T \\ H, T, T, H \\ T, H, H, T \\ T, H, T, H \\ T, T, H, H \\
\hline
T, H, H, H \\ H, T, H, H \\ H, H, T, H \\ H, H, H, T \\
\hline
H, H, H, H
\end{Bmatrix} &&\text{becomes} &&X = \begin{Bmatrix} 0 \\1 \\ 2 \\ 3 \\ 4 \end{Bmatrix}
\end{align}
$$

There is 1 way of getting $X = 0$ , 4 ways of getting $X = 1$, 6 ways of getting $X = 2$, 4 ways of getting $X = 3$, and 1 way of getting $X = 4$.

What is the probability that $X = 2$?

$$
\Pr(X = 2) = \frac{6}{16} = 0.375
$$

```{r}
dbinom(x = 2, size = 4, prob = 1/2)
```

Use the correct R function to calculate the probability that $X \leq 1$ ?

$$
\Pr(X \leq 1) = \Pr(X = 0) + \Pr(X = 1) = \frac{1 + 4}{16} = 0.3125
$$

```{r}
pbinom(1, size = 4, prob = 1/2)
```

### Exercise

::: callout-note
Use `rbinom()` to generate 1000 samples from the Binomial distribution with `size = 5` and `prob = 2/3`. Plot the results in `ggplot` using `geom_bar`.

Estimate the probability that $X$ is an even number using the **`mean()`** function on those 1000 values.
:::

```{r}
x <- rbinom(1e3, size = 5, prob = 2/3)

tibble(x = as_factor(x)) |> 
  ggplot(aes(x)) + 
  geom_bar(width = 1/2)

mean(x == 2 | x == 4)
```

```{r}
sum(dbinom(c(2, 4), size = 5, prob = 2/3))
```

*Note. You can get sophisticated in finding even numbers by using the "modulus" operator.*

```{r}
mean(x %% 2 == 0)
```

*Actually, I just discovered that [zero is technically an even number](https://www.bbc.com/news/magazine-20559052)* 🤯

```{r}
mean(x == 2 | x == 4 | x == 0)
```

### Exercise

::: callout-note
Revisit what you did @sec-four-coins-1 and @sec-four-coins-2.

-   What is the probability that $X = 2$ when $p = 1/3$ and $n = 4$?

-   Verify this by using the correct R function to calculate the probability that $X = 2$?
:::

$$
\Pr\bigg(X = 2 \mid p =\frac{1}{3}, n =4 \bigg) = 6 \times \bigg(\frac{1}{3}\bigg)^2 \times \bigg(\frac{2}{3}\bigg)^2 = 0.2962963
$$

```{r}
dbinom(x = 2, size = 4, prob = 1/3)
```

### Exercise

::: callout-note
Suppose that $p = 2/3$ and $n = 15$

-   Use `dbinom()` to verify the results in @eq-binomial-pmf.

-   Use `pbinom()` to verify the results in @eq-binomial-cdf.
:::

`dbinom()`

```{r}
x <- 0:15
choose(15, x) * (2/3)^x * (1/3)^(15 - x)
dbinom(x, size = 15, prob = 2/3)
```

`pbinom()`

```{r}
cumsum(choose(15, x) * (2/3)^x * (1/3)^(15 - x))
pbinom(x, size = 15, prob = 2/3)
```

## Likelihood

### Exercise

```{r}
grid <- tibble(prob = seq(0, 1, by = 0.001))
grid$like <- dbinom(21, 47, grid$prob)

grid <- grid |> 
  mutate(clike_raw = cumsum(like)) |> 
  mutate(clike_normalized = clike_raw / sum(like)) 

grid |> 
  filter(prob == 0.588)
```

The value of `clike_normalized` gives the chances that `prob` is equal to or less than 0.588. To answer the question I just need to subtract 1 from that.

```{r}
1 - 0.9752575
```

```{r}
grid |> 
  ggplot(aes(prob, clike_normalized)) + 
  geom_line() + 
  geom_vline(xintercept = 0.588, linetype = "dashed") +
  geom_hline(yintercept = 0.9752575, color = "steelblue1")
```
