---
title: "Week 9"
callout-appearance: simple
callout-icon: false
---

This week we will review the idea of ***reconstructing*** $2 \times 2$ and $2 \times 2 \times 2$ tables using statistical models. This should be a straightforward repeat of the things we did on Tuesday. Don't forget to look at [Steve's slides](https://github.com/vaiseys/soc-stats-1/tree/main/old_slides) for a refresher.

```{r}
#| code-summary: "Set up"
#| message: false

library(tidyverse)
library(gssr)
theme_set(theme_light(base_family = "Optima"))

gss18 <- gss_get_yr(2018) 

d <- gss18 |> 
  select(attend, polviews, cappun) |> 
  haven::zap_missing() |> 
  haven::zap_labels() |> 
  mutate(
    weekly = if_else(attend >= 7, 1L, 0L),
    conservative = if_else(polviews >= 5, 1L, 0L),
    cappun = if_else(cappun == 1, 1L, 0L)
    ) |> 
  drop_na() |> 
  select(cappun, weekly, conservative) 
```

## $2 \times 2 \times 2$

### Exercise

The following contingency table shows the association between 2 variables:

-   `cappun`: favor or oppose death penalty for murder

-   `weekly`: attends religious ceremonies one or more times a week

```{r}
crosstab1 <- table(cp = d$cappun, weekly = d$weekly)
crosstab1
## column percentages
prop.table(crosstab1, margin = 2)
```

::: callout-note
**Exercise:**

-   Express the relationship between `cappun` ( $Y$ ) and `weekly` ( $X$ ) as a linear probability model.

-   Express the relationship between `cappun` ( $Y$ ) and `weekly` ( $X$ ) as a logistic regression model.
:::

::: callout-important
**Bonus:**

Calculate the standard error of $\beta$ (in the linear probability model). What is the 95% confidence interval?
:::

### Exercise

The following table adds a third variable to the previous contingency table:

-   `conservative`: identifies as either "slightly conservative", "conservative", or "extremely conservative."

```{r}
crosstab2 <- table(cp = d$cappun, weekly = d$weekly, conservative = d$conservative)
crosstab2
## column percentages
prop.table(crosstab2, margin = c(2, 3))
```

::: callout-note
**Exercise:**

-   Express the relationship between `cappun` ( $Y$ ), `weekly` ( $X_1$ ), and `conservative` ( $X_2$ ) as a linear probability model.

-   Express the relationship between cappun ( $Y$ ), weekly ( $X_1$ ), and conservative ( $X_2$ ) as a logistic regression model.
:::

::: callout-tip
Hint: You might find it easier to express this $2 \times 2 \times 2$ table so that it corresponds to Steve's slides.

```{r}
as_tibble(crosstab2) |> 
  mutate(cp = if_else(cp == "1", "favors", "opposes")) |> 
  pivot_wider(names_from = cp, values_from = n)
```
:::

::: callout-important
**Bonus:**

Calculate the standard error of $\beta_\text{weekly}$ (in the linear probability model). What is the 95% confidence interval?
:::

## Exercise

::: callout-note
Look at the GSS data for 2018.

-   Repeat the $2 \times 2 \times 2$ table exercise for three variables of your choosing.
:::

## Exercise

::: callout-note
Look at the GSS data for 2018.

-   Repeat the $2 \times 2 \times 2$ table exercise for three variables of your choosing.
:::
