---
title: "Week 5"
callout-appearance: simple
callout-icon: false
---

The purpose of this week's homework is to fully understand the binomial probability distribution. We will build up to it from scratch using some 🌟 math 🌟 . We will then revisit some of the things we've done in class.

R has four built-in forms of working with probability distributions.

See `?distributions` for a list of common distributions contained in R.

To work with a **binomial distribution** we have the following functions:

-   `dbinom()`: probability (mass)[^week5-1] function

-   `pbinom()`: cumulative probability distribution

-   `rbinom()`: draw `n` samples from `dbinom()`

-   `qbinom()`: quantile distribution; this is the *inverse* of `pbinom()`.

    What does "inverse" mean here? We'll get to that, but it basically means that---all else equal---the output of input to `pbinom()` corresponds to the output of `qbinom()` (and vice versa).

    ```{r}
    input <- 350
    output <- pbinom(input, size = 1000, prob = 0.345623)
    qbinom(output, size = 1000, prob = 0.345623)
    ```

[^week5-1]: We will get to the meaning of this "mass" word here in a future exercise when we work with *continuous* random variables. A *continuous* random can take an infinite number of possible values. Since the number of outcomes is *uncountable*, we need to take a different approach to how we compute its probability distribution. For now, we can safely ignore this technicality.

## Probability Primer

*Keep track of the **bold words***.

A **random variable** assigns some number to each outcome of a random **event**.

Suppose we flip a coin exactly one time. Then the set of all outcomes---or **sample space---**is given by the following *set*:

$$
\Omega = \{H, T\}
$$

<aside>An **event** is a well-defined *subset* of $\Omega$.</aside>

If we assume that the coin is fair---i.e., that both events are equally likely---then we get the following probabilities.

$$
\begin{align}
\Pr(H) = \frac{1}{2}, && \Pr(T) = \frac{1}{2}
\end{align}
$$

A **probability measure** assigns a number between 0 and 1 to every event in the sample space ( $\Omega$ ). Moreover, the sum of all probabilities must equal one. *Those are the rules!*

Going back to the notion of a **random variable**, if we are recording the number of "heads," then $H$ gets assigned 1 and $T$ gets assigned to 0.

$$
\begin{align}
X\Big(\{H\}\Big) = 1, && X\Big(\{T\}\Big) = 0
\end{align}
$$

## Bernoulli distribution

These are the building blocks of what's known as a *Bernoulli* **probability distribution***,* which takes the following form:

$$
\Pr(X = x) = \begin{cases} 
    p &\text{if} \ x = 1 \\\\
    1 - p &\text{if} \ x = 0 \\\\
    0 &\text{if} \ x = \text{anything else}
\end{cases}
$$ {#eq-bernouilli-pmf}

*Note. If we flip a fair coin, we are basically saying that* $X$ *will follow a Bernouilli distribution with* $p = 0.5$*.*

::: callout-tip
You should be able to define in your own words the following terms: **random variable**, **sample space**, and **event.**
:::

The **cumulative distribution function** returns the probability that an outcome is *less than or equal* to $x$. It looks exactly like this:

$$
\Pr(X \leq x) = \begin{cases} 
    0 &\text{if} \ x < 0 \\\\
    1 - p &\text{if} \ x = 0 \\\\
    1 &\text{if} \ x = 1 \\\\
    1 &\text{if} \ x > 1
\end{cases}
$$ {#eq-bernouilli-cdf}

### Exercise

The Bernoulli distribution is a special case of the **binomial distribution**. If we set the argument `size = 1`, we basically have a Bernouilli distribution.

::: callout-note
Suppose that $p = 2/3$.

-   Use `dbinom()` to verify the results in @eq-bernouilli-pmf.

-   Use `pbinom()` to verify the results in @eq-bernouilli-cdf.

-   Use `rbinom()` to generate 100 samples from the Bernouilli distribution with `prob = 2/3`. Plot the results in `ggplot` using `geom_bar`.
:::

## Binomial distribution

The binomial distribution comes from adding $n$ Bernouilli distributions. Thus, it has two parameters: $n$ and $p$. This "$n$" corresponds to the `size` argument in the R functions we've been working with.

### Exercise {#sec-four-coins-1}

Suppose we flip a fair coin exactly four times.

::: callout-note
Draw the sample space $\Omega$ for this "experiment."

*Note. You can* *pencil and paper and insert a photo of the image if you want to. But I encourage you to play around with the* $\LaTeX$ *formulas.*
:::

::: callout-tip
Hint: There's this little thing in combinatorics called the *multiplication rule.* The number of ways in which we can flip four coins is $2 \times 2 \times 2 \times 2 = 16$. Thus, your sample space should consist of 16 events.
:::

### Exercise {#sec-four-coins-2}

Turn the previous sample space into a random variable that counts the number of heads ( $H$ ) in each event.

::: callout-note
How many possible ways are there to get $X = 0$, $X = 1$, $X = 2$, $X = 3$, and $X = 4$?

What is the probability that $X = 2$?

Verify this by using the correct R function to calculate the probability that $X = 2$?

Use the correct R function to calculate the probability that $X \leq 1$?
:::

### Exercise

::: callout-note
Use `rbinom()` to generate 1000 samples from the Binomial distribution with `size = 5` and `prob = 2/3`. Plot the results in `ggplot` using `geom_bar`.

Estimate the probability that $X$ is an even number using the **`mean()`** function on those 1000 values.
:::

::: callout-tip
You can calculate (almost) the same number by adding up the following values:

```{r}
dbinom(c(2, 4), size = 5, prob = 2/3)
```
:::

### Unfair Coins and Independence

So far, you've seen that it's easy to calculate probabilities by hand when we have fair coins---i.e., when $p=1/2$. It's just addition and division.

But what happens when $p \neq 0.5$?

If we assume that these coin flips are **independent**---i.e., that getting $H$ during the first flip doesn't affect the chances of getting $H$ during the second flip---then we have a very convenient of calculating *joint probabilities.*

We say that two events $A$ and $B$ are **independent** if and only if $\Pr(A \cap B) = \Pr(A) \times \Pr(B)$.

In terms of fair coin flips, we would have that the probability of getting a sequence of $HTHH$ is the same as:

$$
\Pr(HTHH) = \Pr(H)\Pr(T)\Pr(H)\Pr(H) = \frac{1}{16}
$$

This same procedure holds if the coin is unfair.

For example:

If $\Pr(H) = \frac{1}{3}$, then

$$
\Pr(HTHH) = \frac{1}{3} \cdot \frac{2}{3} \cdot \frac{1}{3} \cdot \frac{1}{3} = \frac{2}{81}
$$

### Exercise

::: callout-note
Revisit what you did @sec-four-coins-1 and @sec-four-coins-2.

-   What is the probability that $X = 2$ when $p = 1/3$ and $n = 4$?

-   Verify this by using the correct R function to calculate the probability that $X = 2$?
:::

### Binomial distribution, the formula

Looking back at what you've done, you should be able to understand that the formula for a probability distribution with $n=4$ has a sample space with 16 events, and that there are five possible outcomes $X = \{0, 1, 2, 3, 4 \}$.

The **Binomial** **probability distribution**, with parameters $p$ and $n = 4$ takes the following form:

$$
\Pr(X = x) = \begin{cases} 
    1 \ (1 - p)^4 &\text{if} \ x = 0 \\
    4 \ p(1 - p)^3 &\text{if} \ x = 1 \\ 
    6 \ p^2(1-p)^2 &\text{if} \ x = 2 \\
    4 \ p^1(1-p)^3 &\text{if} \ x = 3 \\
    1 \ p^4 &\text{if} \ x = 4 \\
    0 &\text{if} \ x = \text{anything else}
\end{cases}
$$

The 1s, 4s, and and 6s come from counting the number of ways in which $X$ can equal one of these numbers. You should have gotten these results in @sec-four-coins-2. If you didn't, go back because you did something wrong!

However, this is not how you'll see binomial distributions written out in the wild. We need new notation in order to write **any** binomial distribution, which we get by using the so-called binomial coefficient:

$$
{n \choose x} = \frac{n!}{x! (n - x)!}
$$

<aside>You might remember some of this from the GRE.</aside>

So the probability (mass) function of **any** *binomial distribution* is then this:

$$
\Pr(X = x) = {n \choose x} p^x (1-p)^{n-x}
$$ {#eq-binomial-pmf}

And the cumulative distribution function is as follows:

$$
\Pr(X \leq x) = \sum_{i = 0}^x {n \choose x} p^x (1-p)^{n-x}
$$ {#eq-binomial-cdf}

### Exercise

::: callout-note
Suppose that $p = 2/3$ and $n = 15$

-   Use `dbinom()` to verify the results in @eq-binomial-pmf.

-   Use `pbinom()` to verify the results in @eq-binomial-cdf.
:::

::: callout-tip
Hint: You should strive to verify these results using R. There's a function called `choose()` which calculates the binomial coefficient.
:::

::: callout-tip
Hint: Note that because `pbinom` is just adding different pieces of `dbinom` together, the following two lines of code are equivalent:

```{r}
pbinom(q = seq(-1, 6), size = 5, prob = 0.75)
cumsum(dbinom(x = seq(-1, 6), size = 5, prob = 0.75))
```
:::

## Likelihoods

Likelihood are similar to a probabilities, except that we think of them as a function of parameters ( $X$ is "fixed"); in the previous exercises, we assumed fixed parameters and tried to answer questions about $X$. Now we do the opposite.

This introduces one weird implication:

Suppose we have a binomial distribution with $p = 0.5$ and $n = 55$. Then, summing over all possible values of $X$ should add up to $1$.

```{r}
x <- 0:55
sum(dbinom(x, size = 55, prob = 1/2))
```

This *does not* happen if we take the same formula as a function of the parameter $p$ with $X = 22$.

```{r}
p <- seq(0, 1, by = 0.001)
sum(dbinom(x = 22, size = 55, prob = p))
```

Furthermore, we didn't even sum over all the possible values of $p$. The vector `p` (in this example) is of length 1001, but $p$ can actually take an infinite range of values! This is why Steve did the the whole "normalizing" thing in last [week's class](https://github.com/vaiseys/soc-stats-1/blob/main/demos/day-10.R) in order to turn the likelihoods into a *probability distribution* for $p$.

> *---What? Did you just give probability* $p$ it's own probability distribution?
>
> ---Yes.

### Exercise

Let's revisit Steve's grid approximation code.

I'll give you a start.

```{r}
#| message: false

library(tidyverse)
grid <- tibble(prob = seq(0, 1, by = 0.001))
grid$like <- dbinom(21, 47, grid$prob) ## MAKE SURE YOU USE THESE VALUES
```

::: callout-note
Following Steve's code, calculate `clike_raw` and `clike_normalized`.

Looking at this data frame, what are the chances that `prob` is equal to or greater than 0.588?
:::

<aside>

Modified on 2023-10-01.

This question used to read "what probability would you assign to `prob` having a value equal to or greater than 0.588?". Hopefully the wording is easier to understand.

</aside>

::: callout-tip
Hint: You might want to look at something like this:

```{r}
#| eval: false
grid |> filter(prob == 0.588) 
```
:::
